---
output: github_document
bibliography: inst/references.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# The galamm package for fitting generalized additive latent and mixed models

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/LCBC-UiO/galamm/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/LCBC-UiO/galamm/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

galamm is an R package for fitting multivariate generalized additive mixed models with factor structures. The package is currently under development, and changes to the API must be expected. 

The package uses an interface similar to [PLmixed](https://cran.r-project.org/package=PLmixed), which fits generalized linear mixed models with factor structures. However, galamm allows semiparametric estimations with [gamm4](https://cran.r-project.org/package=gamm4), and hence more flexibly allows estimation of nonlinear effects. If you don't need semiparametric estimation, on the other hand, your should use definitely [PLmixed](https://cran.r-project.org/package=PLmixed), as it is a much more mature package.

The plan for the package is to provide tool for fast and scalable estimation of generalized linear latent and mixed models (GLLAMMs) [@rabe-heskethGeneralizedMultilevelStructural2004] and their semiparametric extensions which we have termed generalized additive latent and mixed models (GALAMMs) [@sorensenLongitudinalModelingAgedependent2021]. This is already possible to some extent, using the function `marginal_likelihood`, as is documented in the vignette on maximum likelihood estimation. To obtain this, we combine sparse matrix computations with RcppEigen [@batesFastElegantNumerical2013] and automatic differentiation using the C++ library [autodiff](https://autodiff.github.io/) [@lealAutodiffModernFast2018].

## Installation

You can install the development version of galamm from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("LCBC-UiO/galamm")
```
## Example

Here is a basic example with three items measuring an underlying latent trait.

```{r example, eval=FALSE}
library(galamm)

# Define factor loading matrix
load.mat <- matrix(c(1, NA, NA), ncol = 1)
dimnames(load.mat) <- list(c("item1", "item2", "item3"), NULL)

# Fit a galamm
mod <- galamm(
  formula = value ~ s(x, by = weight),
  random = ~(1|id),
  data = dat1,
  load.var = "item",
  lambda = load.mat,
  factor = "weight",
  optim_control = list(trace = 3)
  )
```

## References
