---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# The galamm package for fitting generalized additive latent and mixed models

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/LCBC-UiO/galamm/workflows/R-CMD-check/badge.svg)](https://github.com/LCBC-UiO/galamm/actions)
<!-- badges: end -->

galamm is an R package for fitting multivariate generalized additive mixed models with factor structures. The package is currently under development, and changes to the API must be expected. If you are looking for a package for fitting generalized linear mixed models with factor structures, check out [PLmixed](https://cran.r-project.org/package=PLmixed).

## Installation

You can install the development version of galamm from [GitHub](https://github.com/) with:

``` r
# install.packages("remotes")
remotes::install_github("LCBC-UiO/galamm")
```
## Example

Here is a basic example with three items measuring an underlying latent trait.

```{r example}
library(galamm)

# Define factor loading matrix
load.mat <- matrix(c(1, NA, NA), ncol = 1)
dimnames(load.mat) <- list(c("item1", "item2", "item3"), NULL)

# Fit a galamm
mod <- galamm(
  formula = value ~ s(x, by = weight),
  random = ~(1|id),
  data = dat1,
  load.var = "item",
  lambda = load.mat,
  factor = "weight",
  optim_control = list(trace = 3)
  )
```

