---
title: "Introduction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: ../inst/references.bib
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(galamm)
```

The `galamm` packages estimates generalized additive latent and mixed models (GALAMMs). GALAMMs are described in @sorensenLongitudinalModelingAgeDependent2023 as an extension of generalized linear latent and mixed models (GLLAMMs) [@rabe-heskethGeneralizedMultilevelStructural2004]. 

In this vignette we demonstrate basic model formulation.

```{r}
head(dat1)
```

The dataset illustrates measurements of a latent response variable. The variables are

- `id`: an identifier for the study subjects.
- `item`: an identifier for one of three items which measure an underlying latent trait of interest.
- `y`: the measurement of the given item.
- `x`: a predictor variable.

The plot below shows the measurements of each item.

```{r, fig.width=4}
sdat <- split(dat1, f = dat1$item)
for(i in seq_along(sdat)){
  plot(sdat[[i]]$x, sdat[[i]]$y, col = i + 1,
       xlab = "x", ylab = "y", 
       main = paste("Item", i))
}
```



### Model definition

Letting $y_{ij}$ denotes the $i$th measurement of the $j$th subject, we assume the measurement part

$$y_{ij} = \beta_{0} + \eta_{j} \boldsymbol{\lambda}^{T} \boldsymbol{\delta}_{ij}  + \epsilon_{ij}$$

where $\boldsymbol{\delta}_{ij}$ is a vector whose $k$th element equals 1 if the $i$th measurement of the $j$th subject is a measurement of item $k$, and $\boldsymbol{\lambda} = (1, \lambda_{2}, \lambda_{3})^{T}$ is a vector of factor loadings, whose first element is set to 1 for identifiability. Furthermore, $\eta_{j}$ denotes a latent variable for subject $j$, assumed identical across all three items, and $\epsilon_{ij} \sim N(0, \sigma^{2})$ is the residual term.

Next, the structural part relates the latent variables to the predictor variables,

$$\eta_{j} = f(x_{j}) + \zeta_{j}$$

where $x_{j}$ is a predictor variable for subject $j$ which does not vary within subjects, and $\zeta_{j} \sim N(0, \psi)$ is a disturbance term (random intercept).

Plugging the structural model into the measurement model yields the reduced form

$$y_{ij} = \beta_{0} + f(x_{j}) \boldsymbol{\lambda}^{T} \boldsymbol{\delta}_{ij} + \zeta_{j} \boldsymbol{\lambda}^{T} \boldsymbol{\delta}_{ij}  + \epsilon_{ij}$$

At fixed values of $\boldsymbol{\lambda}$, the reduced form defines a generalized additive mixed model (GAMM) [@woodGeneralizedAdditiveModels2017a] with varying coefficient term $f(x_{j}) \boldsymbol{\lambda}^{T} \boldsymbol{\delta}_{ij}$ and random effect term $\zeta_{j} \boldsymbol{\lambda}^{T} \boldsymbol{\delta}_{ij}$. Identifying this lets us use profile likelihood estimation similar to what @jeonProfileLikelihoodApproachEstimating2012 proposed for a subset of GLLAMMs, and @rockwoodEstimatingComplexMeasurement2019 implemented in the R package `PLmixed`. The `galamm` package instead maximizes the Laplace approximate marginal likelihood directly, as described in another vignette.



## References
