---
title: "Heteroscedastic Linear Mixed Models"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
bibliography: ../inst/REFERENCES.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Heteroscedastic Linear Mixed Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(galamm)
```

At the moment, galamm supports group-wise heteroscedasticity in Gaussian response models. Referring to the model formulation outlined in the [introductory vignette](https://lcbc-uio.github.io/galamm/articles/galamm.html), the response model and nonlinear predictor can be easily combined in this particular case, to give

$$
y_{i} = \sum_{s=1}^{S} f_{s}\left(\mathbf{x}_{i}\right) + \sum_{l=2}^{L}\sum_{m=1}^{M_{l}} \eta_{m}^{(l)} \mathbf{z}^{(l)}_{im}{}^{'}\boldsymbol{\lambda}_{m}^{(l)} + \epsilon_{g(i)},
$$

where subscript $i$ refers to the $i$th elementary unit of observation, i.e., the $i$th row in the dataframe. $g(i)$ refers to the group to which the $i$th observation belongs, with each grouping having a separately estimated residual variance, $\epsilon_{g} \sim N(0, \sigma_{g}^{2})$.

In the future, we plan to also support other types of residual terms, including autocorrelation and residuals that depend on continuous variables. Such features are currently supported by the R packages [nlme](https://cran.r-project.org/package=nlme) [@pinheiroMixedEffectsModelsSPLUS2000], [mgcv](https://cran.r-project.org/package=mgcv) [@woodGeneralizedAdditiveModels2017a], and [gamlss](https://cran.r-project.org/package=gamlss) [@rigbyGeneralizedAdditiveModels2005], however, of these only nlme provides computationally efficient estimation of mixed effects models with a large number of grouping levels, and only with strictly nested groups. If you are aware of other packages implementing such functionality, please [let us know](https://github.com/LCBC-UiO/galamm/issues).

## Group-Wise Heteroscedasticity

The package includes a simulated dataset `hsced`, in which the residual variance varies between items.

```{r}
head(hsced)
```

We specify the error structure using an additional formula object, `~ (1 | item)`, specifying that a different constraint term should be included per item.

```{r}
mod <- galamm(
  formula = y ~ x + (1 | id),
  weights = ~ (1 | item),
  data = hsced
)
```

The output shows that for item 2, the residual variance is twice that of item 1.

```{r}
summary(mod)
```

We can confirm that the lme function from the nlme package gives the same result. It reports the multiplies on the standard deviation scale, so since $1.412369^2 = 1.995$, the results are identical.

```{r}
library(nlme)
mod_nlme <- lme(y ~ x,
  data = hsced, random = list(id = ~1),
  weights = varIdent(form = ~ 1 | item), method = "ML"
)
summary(mod_nlme)
```


The diagnostic plot also looks good.

```{r}
plot(mod)
```

We can compare the model to one with homoscedastic residuals.

```{r}
mod0 <- galamm(
  formula = y ~ x + (1 | id),
  data = hsced
)
```


Reassuringly, the correct model is chosen in this simple simulated case.

```{r}
anova(mod, mod0)
```



# References
