---
title: "Models with Mixed Response Types"
output: rmarkdown::html_vignette
bibliography: ../inst/references.bib
vignette: >
  %\VignetteIndexEntry{Models with Mixed Response Types}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

```{r setup}
library(galamm)
library(PLmixed)
library(lme4)
```

This vignette describes how `galamm` can be used to estimate models with mixed response types.

## Mixed Normal and Binomial Response

We start with the `mresp` dataset, which comes with the package. The variable "itemgroup" defines the response type; it equals "a" for normally distributed responses and "b" for binomially distributed responses.

```{r}
mixed_resp <- galamm(
  formula = y ~ x + (0 + itemgroup | id),
  data = mresp,
  family = c(gaussian, binomial),
  family_mapping = ifelse(mresp$itemgroup == "a", 1L, 2L)
)

summary(mixed_resp)
```


## Covariate Measurement Error Model

This example is taken from Chapter 14.2 in @skrondalGeneralizedLatentVariable2004, which I refer to for further interpretation and discussion. The main purpose here is to confirm that we reproduce their results, and to show the syntax.

The model is defined as follows. First, true fiber intake for subject $j$ is assumed to depend on age and on whether the person is a bus driver or banking staff, and their interaction.

$$\eta_{j} = \mathbf{x}_{j}'\boldsymbol{\gamma} + \zeta_{j}.$$

For a part of the sample, there are two measurements of fiber intake, and we can hence estimate a measurement model, as follows:

$$y_{ij} = \eta_{j} + \epsilon_{ij}, \qquad \epsilon_{ij} = N(0, \theta)$$

Finally we define a disease model for the probability of coronary heart disease, as

$$\text{logit}[P(D_{j}=1 | \eta_{j})] = \mathbf{x}_{j}'\boldsymbol{\beta} + \lambda \eta_{j}.$$

Stacking the three responses, which are fiber intake at times 1 and 2, and coronary heart disease, we can define the joint model as a GLLAMM with linear predictor

$$\nu_{ij} = d_{3i} \mathbf{x}_{j}'\boldsymbol{\beta} + \mathbf{x}_{j}'\boldsymbol{\gamma}\left[(d_{1i} + d_{2i}) + \lambda d_{3i}\right] + \zeta_{j} \left[(d_{1i} + d_{2i}) + \lambda d_{3i}\right],$$

where $d_{1i}$ and $d_{2i}$ are indicators for fiber measurements at timepoints 1 and 2, and $d_{3i}$ is an indicator for coronary heart disease.

At this point it may help to look at the first few rows of the dataset.

```{r}
head(diet, 10)
```

With this dataset, the response $y$ is normally distributed when $d_{1i}=1$ or $d_{2i}=1$, but binomially distributed when $d_{3i}=1$. These two response processes are connected through the latent variable $\eta_{j}$.


```{r}
lam <- matrix(c(1, 1, NA), ncol = 1)
mod <- galamm(
  # Add loadings in X
  # Add mixed response option.
  formula = y ~ (age * bus):chd + loading : (age * bus) + (0 + loading | id), 
  data = diet,
  family = c(gaussian, binomial),
  family_mapping = ifelse(diet$item %in% 1:2, 1L, 2L),
  factor = list("loading"), 
  load.var = "item", 
  lambda = list(lam)
  )
```



# References
