---
title: "Interactions Between Latent and Observed Variables"
output:
  rmarkdown::html_vignette:
    fig_width: 6
    fig_height: 4
bibliography: ../inst/REFERENCES.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{Interactions Between Latent and Observed Variables}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = ""
)
```

```{r setup}
library(galamm)
```

This vignette describes how `galamm` can be used to model interactions between latent and observed variables. The models described here can be considered extensions of the covariate measurement error model described in the [vignette on mixed response types](https://lcbc-uio.github.io/galamm/articles/mixed_response.html#covariate-measurement-error-model), by allowing the latent variables to interaction with observed variables.

## Linear Mixed Model with Latent Covariates

For this example we use the simulated `latent_covariates` dataset, of which the first six rows are displayed below:

```{r}
head(latent_covariates)
```

### Model Formulation

The response variable `y` contains both measurements of a latent variable and measurements of the response that we actually are interested in modeling, and the `type` variable distinguishes these responses. In this case we have complete observations for each subject ID, and for a given ID, the measurement model can be written as follows:

$$
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3}
\end{pmatrix}
=
\boldsymbol{\beta}_{0} +
\begin{pmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & 1 & x
\end{pmatrix}
\begin{pmatrix}
1 \\
\lambda_{2} \\
\lambda_{3} \\
\lambda_{4}
\end{pmatrix}
\eta 
+
\begin{pmatrix}
0 \\
0 \\
x \beta
\end{pmatrix}
+ \boldsymbol{\epsilon}.
$$

In this equation $\boldsymbol{\beta}_{0} \in \mathbb{R}^{3}$ is a vector of intercepts, $\eta$ is a latent variable, the loading of the latent variable onto the first measurement $y_{1}$ is fixed to 1 for identifiability, $\lambda_{2}$ is the loading of the latent variable onto the second measurement $y_{2}$, $\lambda_{3}$ is the main effect of the latent variable on the response of interest $y_{3}$, $\beta$ is the effect of the observed covariate $x$ on $y_{3}$, and $\lambda_{4}$ is the interaction effect of $x$ and $\eta$ on $y_{3}$. We assume that the residuals $\boldsymbol{\epsilon}$ are independently and identically normally distributed; this assumption is valid in this simulated case, but note that since the response $y_{3}$ is qualitatively different from the measurements $y_{1}$ and $y_{2}$, this assumption will in general not hold, and a [heteroscedastic measurement model](https://lcbc-uio.github.io/galamm/articles/lmm_heteroscedastic.html) should be used, or a [model with mixed response types](https://lcbc-uio.github.io/galamm/articles/mixed_response.html). For a more detailed explanation of this way of formulating latent variable models in matrix form we refer to the first four pages of @rockwoodEstimatingComplexMeasurement2019.

The structural model is simply $\eta = \zeta \sim N(0, \psi)$, where $\psi$ is its variance.

### Model Without Interaction

It can be instructive to start by considering a model in which we fix $\\lambda_{4} = 0$. This type of model would be estimated with the following code:

```{r}
mod0 <- galamm(
  formula = y ~ x:response + (0 + loading | id),
  data = latent_covariates,
  load.var = "type",
  lambda = list(matrix(c(1, NA, NA), ncol = 1)),
  factor = list("loading")
)
```

In the data generating simulations, the true values were $\lambda_{1}=1$, $\lambda_{2} = 1.3$ and $\lambda_{3} = -0.3$. The former two are very well recovered, but the latter is too positive, which is likely due to us omitting the interaction $\lambda_{4}$, whose true value was 0.2.

```{r}
summary(mod0)
```


### Getting Ready for Computation

The measurement model can be equivalently written as

$$
\begin{pmatrix}
y_{1} \\
y_{2} \\
y_{3}
\end{pmatrix}
=
\boldsymbol{\beta}_{0} +
\begin{pmatrix}
1 \\
\lambda_{2} \\
\lambda_{3} + \lambda_{4} x
\end{pmatrix}
\eta 
+
\begin{pmatrix}
0 \\
0 \\
x \beta
\end{pmatrix}
+ \boldsymbol{\epsilon}.
$$

# References
